Abstraction and Invariance for Algebraically Indexed Types

### 2012-07-09 paper/section-general-framework.tex

    \inferrule* [right=TyEx]
    {\Delta, i \mathord: s \vdash A \isType}
    {\Delta \vdash \exists i \mathord: s. A \isType}
  \end{mathpar}}
  \caption{Index expressions and types}
  \label{fig:indexes-and-types}
\end{figure*}
%Given the above data, 
The rules in \autoref{fig:indexes-and-types}
generate two judgements: well-sorted index expressions $\Delta \vdash
e : s$ and well-indexed types $\Delta \vdash A \isType$. Since index
variables may appear in types, types are judged to be well-indexed
with respect to an index context $\Delta$. The rules for well-sorted
index expressions are particularly simple. %: either an index expression
%is a variable that appears in the context (rule \TirName{IVar}), or it
%is an application of an index operation taken from $\IndexOpSet$ to
%other index expressions (rule \TirName{IOp}). 
The rules for well-indexed types include the usual ones
%rules 
for
%constructing types of 
the simply-typed $\lambda$-calculus with unit, sum and tuple types
(rules \TirName{TyUnit}, \TirName{TyArr}, \TirName{TyTuple} and
\TirName{TySum}). The rule \TirName{TyPrim} %allows us to 
forms, from a primitive type $\tyPrimNm{X}$ and appropriately sorted
index expressions $e_1,...,e_n$, the well-indexed type
$\tyPrim{X}{e_1,...,e_n}$. The rule \TirName{TyForall} 
%permits the formation of 
forms universally quantified types, where the universal
quantification ranges over all index expressions of some
sort. Existential types, %are constructed 
formed using the \TirName{TyEx} rule,
%and 
allow for abstraction by hiding. %FIXME: think about the
%introduction of existentials more, perhaps forward ref to their use

@@

\begin{example}[Two-Dimensional Geometry]
  Example~\autoref{ex:two-dim-geo-axioms}
%  The two-dimensional geometry instantiation of the general framework
  uses the assignment $\tyPrimSem{\tyPrimNm{vec}} = \mathbb{R}^2$ and
  $\tyPrimSem{\tyPrimNm{real}} = \mathbb{R}$.
\end{example}

@@

The definition of relational environments is the most subtle part of
our framework. The straightforward approach is to assume an
interpretation of indexing expressions in which sorts are
%interpreted as 
sets and index operations are functions, and then define the
relational interpretation of a primtive type $\tyPrim{X}{e_1,...,e_n}$
in terms of the interpretations of $e_1,...,e_n$. This approach
suffices to prove free theorems and type isomorphisms for
instantiations of our general framework. However, to prove
non-definability results, like the one in
\exref{ex:uninhabited-type}, we need a more refined interpretation
that accounts more precisely for the definable programs permitted by
the algebraic indexing. In this section, we identify the 
properties we require of relational environments in order for
the abstraction theorem to hold. In \autoref{sec:constr-rel-env}, we
give a general construction of relational environments and use it 
to prove several non-definability results.

A relational environment for a context $\Delta$ is a %(dependently typed)
function $\rho$ that, for each primitive type $\tyPrimNm{X}$ and
instantiation of its arguments $\sigma$, assigns a binary relation
$\rho\ \tyPrimNm{X}\ \sigma$ on the index-erasure interpretation of
$\tyPrimNm{X}$:
\begin{displaymath}
  \rho : (\tyPrimNm{X} \in \PrimTypeSet) \to (\Delta \Rightarrow \primTyArity(\tyPrimNm{X})) \to \Rel(\tyPrimSem{\tyPrimNm{X}})
\end{displaymath}
%Every relation environment must respect index expression equality: for
%any $\tyPrimNm{X} \in \PrimTypeSet$ and pair of simultaneous
%substitutions $\Delta \vdash \sigma \Rightarrow
%\primTyArity(\tyPrimNm{X})$ and $\Delta \vdash \sigma' \Rightarrow
%\primTyArity(\tyPrimNm{X})$, if $\Delta \vdash \sigma \equiv \sigma'
%\Rightarrow \primTyArity(\tyPrimNm{X})$, then $\rho\ \tyPrimNm{X}\
%\sigma = \rho\ \tyPrimNm{X}\ \sigma'$.
Relation environments must respect index expression equality: for
all $\tyPrimNm{X} \in \PrimTypeSet$,
$\Delta \vdash \sigma \Rightarrow
\primTyArity(\tyPrimNm{X})$, and $\Delta \vdash \sigma' \Rightarrow
\primTyArity(\tyPrimNm{X})$, if $\Delta \vdash \sigma \equiv \sigma'
\Rightarrow \primTyArity(\tyPrimNm{X})$, then $\rho\ \tyPrimNm{X}\
\sigma = \rho\ \tyPrimNm{X}\ \sigma'$.
%
We write $\mathrm{RelEnv}(\Delta)$ for the set of all relation
environments for a context $\Delta$. Given %a relational environment
$\rho \in \mathrm{RelEnv}(\Delta')$ and %a simultaneous substitution
$\Delta \vdash \sigma \Rightarrow \Delta'$, we can derive the composed
relational environment (with an abuse of notation) $\rho \circ \sigma
\in \mathrm{RelEnv}(\Delta)$ as $(\rho \circ \sigma)\ \tyPrimNm{X}\
\sigma' = \rho\ \tyPrimNm{X}\ (\sigma' \circ \sigma)$.

As we stated above, index contexts $\Delta$ are interpreted as certain
%sets of relation environments, i.e.,~
subsets of $\mathrm{RelEnv}(\Delta)$. Depending on the primitive
operations that we assume for our system, we may have different sets
of relational environments that enforce different invariants. We
require that the sets of relational environments are \emph{valid} in
the following sense:

\begin{definition}
  \label{defn:valid-rel-env-family}
  A family of sets of relational environments $\relEnv{E}(\Delta)
  \subseteq \mathrm{RelEnv}(\Delta)$ is \emph{valid} if the following
  two properties are satisfied:
  \begin{enumerate}
  \item Closure under simultaneous substitutions: if $\rho \in
    \relEnv{E}(\Delta)$ and $\Delta' \vdash \sigma \Rightarrow \Delta$,
    then $\rho \circ \sigma \in \relEnv{E}(\Delta')$.
%  \item If we have a pair of relation environments $\rho_1 \in
%    \relEnv{E}(\Delta', i\mathord:s')$ and $\rho_2 \in
%    \relEnv{E}(\Delta)$, along with a simultaneous substitution
%    $\Delta' \vdash \sigma \Rightarrow \Delta$, such that $\rho_1
%    \circ \pi_{i\mathord:s'} = \rho_2 \circ \sigma$, then there exists
%    a $\rho \in \relEnv{E}(\Delta, i\mathord:s')$ such that $\rho
%    \circ \sigma_{i\mathord:s'} = \rho_1$ and $\rho \circ
%    \pi_{i\mathord:s'} = \rho_2$.
  \item If $\rho_1 \in \relEnv{E}(\Delta', i\mathord:s')$ and $\rho_2
    \in \relEnv{E}(\Delta)$, and $\Delta' \vdash \sigma \Rightarrow
    \Delta$ is such that $\rho_1 \circ \pi_{i\mathord:s'} = \rho_2
    \circ \sigma$, then there exists a $\rho \in \relEnv{E}(\Delta,
    i\mathord:s')$ such that $\rho \circ \sigma_{i\mathord:s'} =
    \rho_1$ and $\rho \circ \pi_{i\mathord:s'} = \rho_2$.
    % such that the outer edges of the
    % following diagram commute, for all primitive types $\tyPrimNm{X} \in
    % \mathit{PrimType}$:
    % % FIXME: check that these compositions are the right way round
    % % FIXME: re-do this as a set of equations
    % % FIXME: give a better English description of why this is needed
    % \begin{displaymath}
    %   \xymatrix{
    %   {\Delta \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[r]^(.45){- \circ \pi_{i\mathord:s'}} \ar[d]_{- \circ \sigma}
    %   &
    %   {\Delta,i\mathord:s' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[d]^{- \circ \sigma_{s'}} \ar@/^/[rdd]^{\rho_1 \tyPrimNm{X}}
    %   \\
    %   {\Delta' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[r]^(.45){- \circ \pi_{i\mathord:s'}} \ar@/_/[rrd]_{\rho_2\ \tyPrimNm{X}}
    %   &
    %   {\Delta,i\mathord:s' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar@{.>}[dr]_{\rho\ \tyPrimNm{X}}
    %   \\
    %   &
    %   &
    %   {\Rel(\tyPrimSem{\tyPrimNm{X}})}
    % }
    % \end{displaymath}
    % Then there exists a relation environment $\rho \in
    % \relEnv{E}(\Delta,i\mathord:s')$ (the dotted arrow) such that the
    % two triangles in the bottom right of the diagram commute.
    % FIXME: mention that we already know that the square commutes
  \end{enumerate}
\end{definition}
%
% For the general
% framework, we assume that, for each index context $\Delta$, we are
% given a set of relation environments $\relEnv{E}(\Delta)$. Note that
% this assignment of sets of relation environments to index contexts is
% not assumed to be compositional: it is not necessarily the case that
% the set of relational environments $\relEnv{E}(\Delta_1,\Delta_2)$ is
% defined in terms of $\relEnv{E}(\Delta_1)$ and
% $\relEnv{E}(\Delta_2)$. However, it must satisfy the following two
% properties:
%
\noindent
These conditions ensure that the relational interpretation of types we
define below behaves correctly with respect to simultaneous
substitution of index expressions (see
\lemref{lem:tyeqsubst-relational}, below).

@@

% The following lemma states that the relation interpretation of types
% that we have defined in this section behaves well: the first part of
% the lemma states that two types that are judgmentally equal are given
% equal relational interpretations, and the second part states that
% substitution of index expressions in types can be interpreted by the
% composition of relational environments with simultaneous
% substitutions.
\begin{lemma}\label{lem:tyeqsubst-relational}
  \begin{enumerate}
  \item If $\Delta \vdash A \equiv B \isType$, then for all $\rho \in
    \relEnv{E}(\Delta)$, $\rsem{A}{\relEnv{E}}{\rho} =
    \rsem{B}{\relEnv{E}}{\rho}$;
  \item If $\Delta' \vdash A \isType$ and $\Delta \vdash \sigma
    \Rightarrow \Delta'$, then for all $\rho \in \relEnv{E}(\Delta)$,
    $\rsem{\sigma^*A}{\relEnv{E}}\rho = \rsem{A}{\relEnv{E}}(\rho
    \circ \sigma)$.
  \end{enumerate}
\end{lemma}
\noindent
Note that the equations in both parts of
\lemref{lem:tyeqsubst-relational} are well-typed by virtue of the
corresponding parts of \lemref{lem:tyeqsubst-erasure}.

@@

%The typing rules for our framework define 
Well-typed programs 
are defined
with respect to an index context $\Delta$ and a type context $\Delta \vdash
\Gamma \isCtxt$. The judgment $\Delta; \Gamma \vdash M : A$ is defined
in \autoref{fig:programs}. The equational theory on types is
incorporated into the type system via the rule \TirName{TyEq}, which
allows for a program that has %is judged to have 
type $A$ to also have any
%other 
equal type $B$ as well.
\begin{figure*}[t]
  \centering
  {\tiny
  \begin{mathpar}
    \inferrule* [right=Var]
    {\Delta \vdash \Gamma \isCtxt \\ x : A \in \Gamma}
    {\Delta; \Gamma \vdash x : A}

@@

    \inferrule* [right=Case]
    {\Delta; \Gamma \vdash M : A + B \\
      \Delta; \Gamma, x : A \vdash N_1 : C \\
      \Delta; \Gamma, y : B \vdash N_2 : C}
    {\Delta; \Gamma \vdash \textrm{case}\ M\ \textrm{of}\ \textrm{inl}\ x.N_1; \textrm{inr}\ y.N_2 : C}

@@

    \inferrule* [right=ExUnpack]
    {\Delta; \Gamma \vdash M : \exists i\mathord:s. A \\
      \Delta, i\mathord:s; \pi_{i\mathord:s}^*\Gamma, x : A \vdash N : \pi_{i\mathord:s}^*B}
    {\Delta; \Gamma \vdash \mathrm{let}\langle[i],x\rangle = M\ \mathrm{in}\ N : B}
  \end{mathpar}}

@@

\begin{theorem}[Abstraction]\label{thm:abstraction}
  If $\Delta; \Gamma \vdash M : A$, then for all $\rho \in
  \relEnv{E}(\Delta)$ and $\eta_1, \eta_2 \in \ctxtSem{\Gamma}$ such
  that $(\eta_1, \eta_2) \in \rsem{\Gamma}{\relEnv{E}}\rho$, we have
  $(\tmSem{M}\eta_1, \tmSem{M}\eta_2) \in \rsem{A}{\relEnv{E}}\rho$.
\end{theorem}

We use the relational interpretation to define the notion of semantic
equivalence, which is a sound approximation of contextual equivalence,
as a consequence of \thmref{thm:abstraction}.
\begin{definition}
  Let $\Gamma_{\mathit{ops}}$ and $\eta_{\mathit{ops}}$ be a context
  of primitive operations and its interpretation as in
  \defref{defn:ctxt-equiv}. Let $\relEnv{E}_{\mathit{ops}}$ be a
  family of sets of relational environments such that for all $\Delta$
  and $\rho \in \relEnv{E}_{\mathit{ops}}(\Delta)$,
  $(\eta_{\mathit{ops}}, \eta_{\mathit{ops}}) \in
  \rsem{\Gamma}{\relEnv{E}_{\mathit{ops}}}\rho$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{A General Construction of Relational Environments}
\label{sec:constr-rel-env}

\newcommand{\Free}{\mathrm{Free}}
\newcommand{\semSort}[1]{\llbracket #1 \rrbracket^{\mathcal{S}}}
\newcommand{\semIndexExp}[1]{\llbracket #1 \rrbracket^{\mathcal{I}}}

We now give a general construction of valid families of sets of
relational environments 
%(in the sense of \defref{defn:valid-rel-env-family}) 
that allow us to derive
useful consequences of \thmref{thm:abstraction}, e.g.,
%in particular, 
%we wish to show 
that certain types are uninhabited. To %motivate 
this end, consider the
%following 
type $\forall t\mathord:\SynTransl{2}.\ \tyPrim{vec}{t + t} \to
\tyPrim{vec}{t}$ involving vectors indexed by translations.
% as in \autoref{sec:affine-vector-ops}.
%\begin{displaymath}
%  \forall t\mathord:\SynTransl{2}.\ \tyPrim{vec}{t + t} \to \tyPrim{vec}{t}
%\end{displaymath}
In \exref{ex:uninhabited-type} we claimed that this type is
uninhabited. Intuitively, it is clear why: given the operations on
vectors we presented in \autoref{sec:affine-vector-ops}, there is no
way of cancelling the additional $t$ in the argument type to get the
result type. More generally, our operations on vectors 
only allow us to combine indices according to the group operations. Thus,
if the index of the result type of a function does not lie within the
free abelian group generated by the indices of the input types, then
there is no way to write a program with this %given 
type. We now
develop this idea to generate a family of sets of relational
environments that is valid in the sense of
\defref{defn:valid-rel-env-family}. The key idea is to assign
different relational interpretations to types with indices that lie
outside some fintely generated group of index expressions.

\paragraph{Some Universal Algebra}
We make use of some concepts of universal algebra, which now recall.
A \emph{model} for our indexing expressions and their equality is
defined as follows. To each sort $s \in \SortSet$ we assign a carrier
set $\semSort{s}$. We extend this assignment to index contexts and
other sequences of sorts (e.g., type arities) using the cartesian
product of sets: $\semSort{i_1\mathord:s_1,...,i_n\mathord:s_n} =
\semSort{s_1} \times ... \times \semSort{s_n}$. For each index
operation $\texttt{f} \in \IndexOpSet$ with $\indexOpArity(\texttt{f})
= ([s_1,...,s_n],s)$, we assume a function $\semIndexExp{\texttt{f}} :
\semSort{s_1,...,s_n} \to \semSort{s}$. For each well-sorted index
expression $\Delta \vdash e : s$, we assign a function
$\semIndexExp{e} : \semSort{\Delta} \to \semSort{s}$ by recursion on
the structure of $e$. We further assume that for each axiom $\Delta
\vdash e \stackrel{ax}\equiv e' : s \in \IndexAxiomSet$, we have
$\semIndexExp{e} = \semIndexExp{e'}$. A \emph{model} is a pair
$\mathcal{M} = (\semSort{\cdot}_{\mathcal{M}},
\semIndexExp{\cdot}_{\mathcal{M}})$ 
%such that all the axioms in
satisfying all the axioms in
$\IndexAxiomSet$.
% are satisfied.

Given a pair of models $\mathcal{M}_1 = (\semSort{\cdot}_1,
\semIndexExp{\cdot}_1)$ and $\mathcal{M}_2 = (\semSort{\cdot}_2,
\semIndexExp{\cdot}_2)$, a \emph{homomorphism} $h : \mathcal{M}_1 \to
\mathcal{M}_2$ is a family of functions $\{ h_s : \semSort{s}_1 \to
\semSort{s}_2 \}_{s \in \SortSet}$ such that for each $\texttt{f} \in
\IndexOpSet$, we have $h_s(\semIndexExp{\texttt{f}}_1 (x_1,...,x_n)) =
\semIndexExp{\texttt{f}}_2(h_{s_1}x_1, ..., h_{s_n}x_n)$.

>>>>

    \inferrule* [right=TyEx]
    {\Delta, i \mathord: s \vdash A \isType}
    {\Delta \vdash \exists i \mathord: s. A \isType}
  \end{mathpar}}
  \caption{Index expressions and types}
  \label{fig:indexes-and-types}
\end{figure*}
%Given the above data, 
The rules in \autoref{fig:indexes-and-types}
generate two judgements: well-sorted index expressions $\Delta \vdash
e : s$ and well-indexed types $\Delta \vdash A \isType$. Since index
variables may appear in types, types are judged to be well-indexed
with respect to an index context $\Delta$. The rules for well-sorted
index expressions are particularly simple. %: either an index expression
%is a variable that appears in the context (rule \TirName{IVar}), or it
%is an application of an index operation taken from $\IndexOpSet$ to
%other index expressions (rule \TirName{IOp}). 
The rules for well-indexed types include the usual ones
%rules 
for
%constructing types of 
the simply-typed $\lambda$-calculus with unit, sum and tuple types
(rules \TirName{TyUnit}, \TirName{TyArr}, \TirName{TyTuple} and
\TirName{TySum}). The rule \TirName{TyPrim} %allows us to 
forms, from a primitive type $\tyPrimNm{X}$ and appropriately sorted
index expressions $e_1,...,e_n$, the well-indexed type
$\tyPrim{X}{e_1,...,e_n}$. The rule \TirName{TyForall} 
%permits the formation of 
forms universally quantified types, where the universal
quantification ranges over all index expressions of some
sort. Existential types, %are constructed 
formed using the \TirName{TyEx} rule,
%and 
allow for abstraction by hiding. %FIXME: think about the
%introduction of existentials more, perhaps forward ref to their use
We say that a type is \emph{quantifier-free} if it does not contain
ether universal or existential quantification.

@@

\begin{example}[Two-Dimensional Geometry]
  \exref{ex:two-dim-geo-axioms}
%  The two-dimensional geometry instantiation of the general framework
  uses the assignment $\tyPrimSem{\tyPrimNm{vec}} = \mathbb{R}^2$ and
  $\tyPrimSem{\tyPrimNm{real}} = \mathbb{R}$.
\end{example}

@@

The definition of relational environments is the most subtle part of
our framework. The straightforward approach is to assume an
interpretation of indexing expressions in which sorts are
%interpreted as 
sets and index operations are functions, and then define the
relational interpretation of a primtive type $\tyPrim{X}{e_1,...,e_n}$
in terms of the interpretations of $e_1,...,e_n$. This approach
suffices to prove free theorems and type isomorphisms. However, to
prove non-definability results, like the one in
\exref{ex:uninhabited-type}, we need a more refined interpretation
that accounts for the definable programs permitted by the algebraic
indexing. Here, we identify the properties we require of relational
environments for the abstraction theorem to hold. In
\autoref{sec:constr-rel-env}, we give two generic constructions of
relational environments.

A relational environment for a context $\Delta$ is a %(dependently typed)
function $\rho$ that, for each primitive type $\tyPrimNm{X}$ and
instantiation of its arguments $\sigma$, assigns a binary relation
$\rho\ \tyPrimNm{X}\ \sigma$ on the index-erasure interpretation of
$\tyPrimNm{X}$:
\begin{displaymath}
  \rho : (\tyPrimNm{X} \in \PrimTypeSet) \to (\Delta \Rightarrow \primTyArity(\tyPrimNm{X})) \to \Rel(\tyPrimSem{\tyPrimNm{X}})
\end{displaymath}
%Every relation environment must respect index expression equality: for
%any $\tyPrimNm{X} \in \PrimTypeSet$ and pair of simultaneous
%substitutions $\Delta \vdash \sigma \Rightarrow
%\primTyArity(\tyPrimNm{X})$ and $\Delta \vdash \sigma' \Rightarrow
%\primTyArity(\tyPrimNm{X})$, if $\Delta \vdash \sigma \equiv \sigma'
%\Rightarrow \primTyArity(\tyPrimNm{X})$, then $\rho\ \tyPrimNm{X}\
%\sigma = \rho\ \tyPrimNm{X}\ \sigma'$.
Relation environments must respect index expression equality: for
all $\tyPrimNm{X} \in \PrimTypeSet$,
$\Delta \vdash \sigma \Rightarrow
\primTyArity(\tyPrimNm{X})$, and $\Delta \vdash \sigma' \Rightarrow
\primTyArity(\tyPrimNm{X})$, if $\Delta \vdash \sigma \equiv \sigma'
\Rightarrow \primTyArity(\tyPrimNm{X})$, then $\rho\ \tyPrimNm{X}\
\sigma = \rho\ \tyPrimNm{X}\ \sigma'$.
%
We write $\mathrm{RelEnv}(\Delta)$ for the set of all relation
environments for a context $\Delta$. Given %a relational environment
$\rho \in \mathrm{RelEnv}(\Delta)$ and %a simultaneous substitution
$\Delta \vdash \sigma \Rightarrow \Delta'$, we can derive the composed
relational environment (with an abuse of notation) $\rho \circ \sigma
\in \mathrm{RelEnv}(\Delta')$ as $(\rho \circ \sigma)\ \tyPrimNm{X}\
\sigma' = \rho\ \tyPrimNm{X}\ (\sigma' \circ \sigma)$.

As we stated above, index contexts $\Delta$ are interpreted as certain
%sets of relation environments, i.e.,~
subsets of $\mathrm{RelEnv}(\Delta)$. Depending on the primitive
operations that we assume for our system, we may have different sets
of relational environments that enforce different invariants. 
% We
% require that the sets of relational environments are \emph{valid} in
% the following sense:

\begin{definition}
  \label{defn:valid-rel-env-family}
  A family of sets of relational environments $\relEnv{E}(\Delta)
  \subseteq \mathrm{RelEnv}(\Delta)$ is \emph{substitutive} if these
  two properties are satisfied:
  \begin{enumerate}
  \item Closure under simultaneous substitutions: if $\rho \in
    \relEnv{E}(\Delta)$ and $\Delta' \vdash \sigma \Rightarrow \Delta$,
    then $\rho \circ \sigma \in \relEnv{E}(\Delta')$.
%  \item If we have a pair of relation environments $\rho_1 \in
%    \relEnv{E}(\Delta', i\mathord:s')$ and $\rho_2 \in
%    \relEnv{E}(\Delta)$, along with a simultaneous substitution
%    $\Delta' \vdash \sigma \Rightarrow \Delta$, such that $\rho_1
%    \circ \pi_{i\mathord:s'} = \rho_2 \circ \sigma$, then there exists
%    a $\rho \in \relEnv{E}(\Delta, i\mathord:s')$ such that $\rho
%    \circ \sigma_{i\mathord:s'} = \rho_1$ and $\rho \circ
%    \pi_{i\mathord:s'} = \rho_2$.
  \item If $\rho_1 \in \relEnv{E}(\Delta', i\mathord:s')$ and $\rho_2
    \in \relEnv{E}(\Delta)$, and $\Delta \vdash \sigma \Rightarrow
    \Delta'$ is such that $\rho_1 \circ \pi_{i\mathord:s'} = \rho_2
    \circ \sigma$, then there exists a $\rho \in \relEnv{E}(\Delta,
    i\mathord:s')$ such that $\rho \circ \sigma_{i\mathord:s'} =
    \rho_1$ and $\rho \circ \pi_{i\mathord:s'} = \rho_2$.
    % such that the outer edges of the
    % following diagram commute, for all primitive types $\tyPrimNm{X} \in
    % \mathit{PrimType}$:
    % % FIXME: check that these compositions are the right way round
    % % FIXME: re-do this as a set of equations
    % % FIXME: give a better English description of why this is needed
    % \begin{displaymath}
    %   \xymatrix{
    %   {\Delta \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[r]^(.45){- \circ \pi_{i\mathord:s'}} \ar[d]_{- \circ \sigma}
    %   &
    %   {\Delta,i\mathord:s' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[d]^{- \circ \sigma_{s'}} \ar@/^/[rdd]^{\rho_1 \tyPrimNm{X}}
    %   \\
    %   {\Delta' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[r]^(.45){- \circ \pi_{i\mathord:s'}} \ar@/_/[rrd]_{\rho_2\ \tyPrimNm{X}}
    %   &
    %   {\Delta,i\mathord:s' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar@{.>}[dr]_{\rho\ \tyPrimNm{X}}
    %   \\
    %   &
    %   &
    %   {\Rel(\tyPrimSem{\tyPrimNm{X}})}
    % }
    % \end{displaymath}
    % Then there exists a relation environment $\rho \in
    % \relEnv{E}(\Delta,i\mathord:s')$ (the dotted arrow) such that the
    % two triangles in the bottom right of the diagram commute.
    % FIXME: mention that we already know that the square commutes
  \end{enumerate}
  A family $\relEnv{E}$ is \emph{pre-substitutive} if only the first
  point holds.
\end{definition}
%
% For the general
% framework, we assume that, for each index context $\Delta$, we are
% given a set of relation environments $\relEnv{E}(\Delta)$. Note that
% this assignment of sets of relation environments to index contexts is
% not assumed to be compositional: it is not necessarily the case that
% the set of relational environments $\relEnv{E}(\Delta_1,\Delta_2)$ is
% defined in terms of $\relEnv{E}(\Delta_1)$ and
% $\relEnv{E}(\Delta_2)$. However, it must satisfy the following two
% properties:
%
\noindent
These conditions ensure that the relational interpretation of types we
define below behaves correctly with respect to simultaneous
substitution of index expressions (see
\lemref{lem:tyeqsubst-relational}, below).

@@

% The following lemma states that the relation interpretation of types
% that we have defined in this section behaves well: the first part of
% the lemma states that two types that are judgmentally equal are given
% equal relational interpretations, and the second part states that
% substitution of index expressions in types can be interpreted by the
% composition of relational environments with simultaneous
% substitutions.
\begin{lemma}\label{lem:tyeqsubst-relational}
  \begin{enumerate}
  \item If $\Delta \vdash A \equiv B \isType$, then for all $\rho \in
    \relEnv{E}(\Delta)$, $\rsem{A}{\relEnv{E}}{\rho} =
    \rsem{B}{\relEnv{E}}{\rho}$;
  \item If $\Delta' \vdash A \isType$ and either $\relEnv{E}$ is
    substitutive or $\relEnv{E}$ is pre-substitutive and $A$ is
    quantifier-free, then for all $\Delta \vdash \sigma \Rightarrow
    \Delta'$ and $\rho \in \relEnv{E}(\Delta)$,
    $\rsem{\sigma^*A}{\relEnv{E}}\rho = \rsem{A}{\relEnv{E}}(\rho
    \circ \sigma)$.
  \end{enumerate}
\end{lemma}
\noindent
Note that the equations in both parts of
\lemref{lem:tyeqsubst-relational} are well-typed by virtue of the
corresponding parts of \lemref{lem:tyeqsubst-erasure}.

@@

%The typing rules for our framework define 
Well-typed programs 
are defined
with respect to an index context $\Delta$ and a type context $\Delta \vdash
\Gamma \isCtxt$. The judgment $\Delta; \Gamma \vdash M : A$ is defined
in \autoref{fig:programs}. The equational theory on types is
incorporated into the type system via the rule \TirName{TyEq}, which
allows for a program that has %is judged to have 
type $A$ to also have any
%other 
equal type $B$ as well.
\begin{figure*}[t]
  \centering
  {\small
  \begin{mathpar}
    \inferrule* [right=Var]
    {\Delta \vdash \Gamma \isCtxt \\ x : A \in \Gamma}
    {\Delta; \Gamma \vdash x : A}

@@

    \inferrule* [right=Case]
    {\Delta; \Gamma \vdash M : A + B \\\\
      \Delta; \Gamma, x : A \vdash N_1 : C \\\\
      \Delta; \Gamma, y : B \vdash N_2 : C}
    {\Delta; \Gamma \vdash \textrm{case}\ M\ \textrm{of}\ \textrm{inl}\ x.N_1; \textrm{inr}\ y.N_2 : C}

@@

    \inferrule* [right=ExUnpack]
    {\Delta; \Gamma \vdash M : \exists i\mathord:s. A \\\\
      \Delta, i\mathord:s; \pi_{i\mathord:s}^*\Gamma, x : A \vdash N : \pi_{i\mathord:s}^*B}
    {\Delta; \Gamma \vdash \mathrm{let}\langle[i],x\rangle = M\ \mathrm{in}\ N : B}
  \end{mathpar}}

@@

\begin{theorem}[Abstraction]\label{thm:abstraction}
  Assume $\relEnv{E}$ is substitutive. If $\Delta; \Gamma \vdash M :
  A$, then for all $\rho \in \relEnv{E}(\Delta)$ and $\eta_1, \eta_2
  \in \ctxtSem{\Gamma}$ such that $(\eta_1, \eta_2) \in
  \rsem{\Gamma}{\relEnv{E}}\rho$, we have $(\tmSem{M}\eta_1,
  \tmSem{M}\eta_2) \in \rsem{A}{\relEnv{E}}\rho$.

  If $\relEnv{E}$ is pre-substitutive, then the above restricts to the
  case when $\Gamma$ only contains types of the form $\forall
  \vec{i}\mathord:\vec{s}. B$, where $B$ is quantifier-free, $A$ is
  quantifier-free and $M$ contains no uses of \TirName{UnivAbs},
  \TirName{ExPack} or \TirName{ExUnpack} and all uses of
  \TirName{UnivApp} are fully applied.
\end{theorem}

We refer to programs satisfying the conditions in the second part of
\thmref{thm:abstraction} as being in the \emph{quantifier-free}
fragment of our framework. These conditions are very similar to those
used in the fragment of System F covered by Hindley-Milner type
inference.

We use the relational interpretation to define semantic equivalence,
which is a sound approximation of contextual equivalence, as a
consequence of \thmref{thm:abstraction}.
\begin{definition}
  Let $\Gamma_{\mathit{ops}}$ and $\eta_{\mathit{ops}}$ be a context
  of primitive operations and its interpretation as in
  \defref{defn:ctxt-equiv}. Let $\relEnv{E}_{\mathit{ops}}$ be a
  substitutive family of sets of relational environments such that for
  all $\Delta$ and $\rho \in \relEnv{E}_{\mathit{ops}}(\Delta)$,
  $(\eta_{\mathit{ops}}, \eta_{\mathit{ops}}) \in
  \rsem{\Gamma}{\relEnv{E}_{\mathit{ops}}}\rho$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Construction of Relational Environments}
\label{sec:constr-rel-env}

\newcommand{\Gen}{\mathrm{Gen}}
\newcommand{\Free}{\mathrm{Free}}
\newcommand{\semSort}[1]{\llbracket #1 \rrbracket^{\mathcal{S}}}
\newcommand{\semIndexExp}[1]{\llbracket #1 \rrbracket^{\mathcal{I}}}

We give two constructions of families of sets of relational
environments that allow us to derive useful consequences of
\thmref{thm:abstraction}.

\paragraph{Compositional Relational Environments}
Our compositional relational environments are defined in terms of
models. To each sort $s \in \SortSet$ we assign a carrier set
$\semSort{s}$. We extend this assignment to index contexts and other
sequences of sorts (e.g., type arities) using the cartesian product of
sets: $\semSort{i_1\mathord:s_1,...,i_n\mathord:s_n} = \semSort{s_1}
\times ... \times \semSort{s_n}$. For each index operation $\texttt{f}
\in \IndexOpSet$ with $\indexOpArity(\texttt{f}) = ([s_1,...,s_n],s)$,
we assume a function $\semIndexExp{\texttt{f}} : \semSort{s_1,...,s_n}
\to \semSort{s}$. For each well-sorted index expression $\Delta \vdash
e : s$, we assign a function $\semIndexExp{e} : \semSort{\Delta} \to
\semSort{s}$ by recursion on the structure of $e$. We further assume
that for each axiom $\Delta \vdash e \stackrel{ax}\equiv e' : s \in
\IndexAxiomSet$, we have $\semIndexExp{e} = \semIndexExp{e'}$. A
\emph{model} is a pair $\mathcal{M} = (\semSort{\cdot}_{\mathcal{M}},
\semIndexExp{\cdot}_{\mathcal{M}})$ satisfying all the axioms in
$\IndexAxiomSet$.

Now fix a model $\mathcal{M}$ and assume, for each primitive type
$\tyPrimNm{X}$, a relational interpretation parameterised by elements
from the model: $R_{\tyPrimNm{X}} :
\semSort{\primTyArity(\tyPrimNm{X})}_{\mathcal{M}} \to
\Rel(\tyPrimSem{\tyPrimNm{X}})$. Our family of sets of relational
environments is $\relEnv{E}_{\mathcal{M},R}(\Delta) = \{ \rho^\delta
\sepbar \delta \in \semSort{\Delta}_{\mathcal{M}} \}$, where
$\rho^\delta\ \tyPrimNm{X}\ (e_1,...,e_n) =
R_{\tyPrimNm{X}}(\semIndexExp{e_1}_{\mathcal{M}}\delta, ...,
\semIndexExp{e_n}_{\mathcal{M}}\delta)$.

\begin{theorem}
  The families $\relEnv{E}_{\mathcal{M},R}$ are substitutive.
\end{theorem}
